---
title: "Used-habitat calibration plots"
author: "Marie Gilbertson"
date: "2025-03-03"
knit: (function(inputFile, encoding) {
      out_dir <- "../Markdown_reports";
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_dir=file.path(dirname(inputFile), out_dir))})
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this report, I'll demonstrate adapting the approach described by [Fieberg et al. 2017](https://doi.org/10.1111/ecog.03123) for validating step-selection models: used-habitat calibration (UHC) plots.  

Commenting will at times seem somewhat minimal. This is done to encourage users to refer to the [original code](http://doi.org/10.13020/D6T590) published by Fieberg et al. Their documentation is excellent. **Both the Fieberg et al. publication and code should be reviewed prior to using the code in this report.**    

In the manuscript associated with this report, we fit individual-level integrated step selection functions (iSSFs) for thousands of trajectories. Replicating this for an RMarkdown report is not practical, so I'll demonstrate our approach at a smaller scale.  

We'll start by loading the packages we'll need. I also always set my seed as standard practice (this makes randomization reproducible).

```{r, results='hide', message=FALSE, warning=FALSE}
##### Clear Environment #####
remove(list=ls())


#### set seed ####
set.seed(1654)


#### load libraries ####
library(plyr)
library(dplyr)
library(amt)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(sf)

```



## Load data

Data prepped for iSSF fitting; one individual, same season, different years. Scaling/centering of predictors already done; levels of habitat covariates are identical between training and testing data sets.

```{r}

uhc.demo.dat <- get(load("../Project_data/UHC_demo_data.Rdata"))
train <- uhc.demo.dat$train
test <- uhc.demo.dat$test

print(head(train), width = Inf)
print(head(test), width = Inf)


```


```{r}

uhc_sim.strat <- function (nsims, xmat, stratum, fit_ssf, z) {
  xmat <- xmat[,-1] #remove intercept
  ustrat <- unique(stratum)
  nstrat <- length(ustrat)
  x_sim_choice <- array(NA, dim = c(nsims, nstrat, ncol(z)))
  beta.hats <- MASS::mvrnorm(nsims, coef(fit_ssf), fit_ssf$model$var)
  ntot.test <- nrow(xmat)
  dimxmat <- ncol(xmat)
  tempdat <- as.data.frame(cbind(1:nrow(xmat), stratum, z))
  names(tempdat)[1] <- "inds"
  for (i in 1:nsims) {
    lp.hat.s <- as.matrix(xmat) %*% beta.hats[i, ]
    tempdat$wx.s <- exp(lp.hat.s)
    tempdat.g <- tempdat %>% group_by(stratum)
    x_sim_choice[i, , ] <- z[sample_n(tempdat.g, 1, weight = wx.s)$inds, 
    ]
  }
  return(x_sim_choice)
}

uhc_dens <- function (rand_sims, dat, avail, gridsize = 500) {
  rand_sims <- as.matrix(rand_sims)
  dat <- as.matrix(dat)
  avail <- as.matrix(avail)
  alldat <- c(rand_sims, dat, avail)
  range.x <- c(min(alldat), max(alldat))
  nsims <- nrow(rand_sims)
  densrand <- matrix(NA, nsims, gridsize)
  densdat <- KernSmooth::bkde(as.vector(dat), range.x = range.x, 
                              gridsize = gridsize)
  for (i in 1:nsims) {
    temp <- KernSmooth::bkde(rand_sims[i, ], range.x = range.x, 
                             gridsize = gridsize)
    densrand[i, ] <- temp$y
  }
  densavail <- KernSmooth::bkde(as.vector(avail), range.x = range.x, 
                                gridsize = gridsize)
  out <- list(densdat = densdat, densrand = densrand, densavail = densavail)
  return(out)
}

```



```{r}

### fit model(s) to training data
mod.formula <- as.formula(paste(
    "case_ ~ 
    log_sl_ + cos.ang + 
    poly(eastness, 2, raw = TRUE) + northness + 
    log_fe.dist + 
    flu_end4*ndvi_end + strata(step_id_)",
    collapse = " "
  ))


training.mod <- train %>% fit_issf(mod.formula, model = T)
  
  

### run simulations with test data and training model
## keep only the rows in mdat.test that are also in the design matrix 
## (need to remove rows with missing data)
 
na.cols <- which(apply(test[,c("sl_", "ta_", "flu_end4", "ndvi_end", "eastness", "northness", "fe.dist")], 2, 
                       function(x) sum(is.na(x))) !=0)
na.dat <- test[,c("sl_", "ta_", "flu_end4", "ndvi_end", "eastness", "northness", "fe.dist")][, na.cols]

na.rows <- unique(unlist(apply(na.dat, 2, function(x) which(is.na(x)))))
mdat.test <- test[-c(na.rows),]


design.mat.test.full <- model.matrix(mod.formula, data=mdat.test)
strata.mat <- design.mat.test.full[,grepl("strata", colnames(design.mat.test.full))]
design.mat.test.full <- design.mat.test.full[,!grepl("strata", colnames(design.mat.test.full))]

z.col <- colnames(design.mat.test.full)
z.col <- z.col[!grepl("Intercept|sl_|ang", z.col)]
z <- model.matrix(mod.formula, 
                  data = mdat.test)
z <- z[,which(colnames(z) %in% z.col)]



uniq.strat <- sub(".*=", "", colnames(strata.mat))
reps.strat <- colSums(strata.mat)
strata.vec <- rep(uniq.strat, reps.strat)
strata.vec <- as.numeric(c(rep("1", nrow(z) - length(strata.vec)), strata.vec))
  
xchoice.full <- NA
xchoice.full <- uhc_sim.strat(nsims = 100,
                            xmat = design.mat.test.full, 
                            stratum = strata.vec, 
                            fit_ssf = training.mod,
                            z = z)

  


  
variables <- colnames(z)
v <- which(variables=="log_fe.dist")

denshats.full <- uhc_dens(rand_sims=xchoice.full[,,v],
                          dat=z[mdat.test$case_,v],
                          avail=z[!mdat.test$case_,v], gridsize = 2000)
    

    
xs <- denshats.full$densdat$x

obs.y <- denshats.full$densdat$y
avl.y <- denshats.full$densavail$y

low.s <- apply(denshats.full$densrand, 2, quantile, prob = 0.025, na.rm = TRUE)
up.s <- apply(denshats.full$densrand, 2, quantile, prob = 0.975, na.rm = TRUE)

pol <- st_sfc(st_polygon(list(cbind(c(xs, rev(xs), xs[1]),c(up.s, rev(low.s), up.s[1])))))
obs.line <- st_sfc(st_linestring(matrix(c(xs, obs.y), ncol = 2)))
avl.line <- st_sfc(st_linestring(matrix(c(xs, avl.y), ncol = 2)))


obs.o.length <- st_length(obs.line)
avl.o.length <- st_length(avl.line)

obs.c.line <- st_intersection(obs.line, pol)
avl.c.line <- st_intersection(avl.line, pol)


obs.c.length <- st_length(obs.c.line)
avl.c.length <- st_length(avl.c.line)

if(any(unlist(lapply(obs.c.line[[1]], function(x) any(class(x)=="POINT"))))){
  obs.c.line <- obs.c.line[[1]][-which(unlist(lapply(obs.c.line[[1]], function(x) any(class(x)=="POINT"))))]
  obs.c.line <- st_sfc(obs.c.line)
}
if(any(unlist(lapply(avl.c.line[[1]], function(x) any(class(x)=="POINT"))))){
  avl.c.line <- avl.c.line[[1]][-which(unlist(lapply(avl.c.line[[1]], function(x) any(class(x)=="POINT"))))]
  avl.c.line <- st_sfc(avl.c.line)
}

obs.prop <- obs.c.length/obs.o.length
avl.prop <- avl.c.length/avl.o.length


pol.df <- data.frame(st_coordinates(pol))
obs.line.df <- data.frame(st_coordinates(obs.line))
avl.line.df <- data.frame(st_coordinates(avl.line))
obs.c.line.df <- data.frame(st_coordinates(obs.c.line))
avl.c.line.df <- data.frame(st_coordinates(avl.c.line))

  
ggplot() + 
  geom_polygon(data = pol.df, aes(x = X, y = Y), fill = "grey") +
  geom_path(data = obs.line.df, aes(x = X, y = Y), linetype = "dashed") + 
  geom_path(data = avl.line.df, aes(x = X, y = Y), color = "red", linetype = "dashed") +
  geom_path(data = obs.c.line.df, aes(x = X, y = Y, group = L1)) + geom_path(data = avl.c.line.df, aes(x = X, y = Y, group = L1), color = "red") +
  ggtitle(paste0("Log(Distance to forest edge)", ": ", "observed in = ", round(obs.prop, 2), "; available in = ", round(avl.prop, 2))) +
  theme_bw() +
  xlab("Scaled log(Distance to forest edge)") + ylab("Density")
  
   

```